EEG2fMRI: Cross-Modal Synthesis for Functional Neuroimaging

### 2020-11-21 article/article.tex
Update EEG-fMRI

Neural activity is central to investigating cognition. In cognitive neuroscience, it has been employed to study the neural correlates of numerous cognitive functions, such as perception, attention, and memory. In cognitive modelling, neural activity has been used to inform the development of a wide range of cognitive architectures and models. In clinical neuroscience, it has been employed for characterizing, diagnosing, and ameliorating a wide range of neurological conditions.

The increased relevance of neural activity in cognitive science prompted the development of a large number of functional neuroimaging techniques. These techniques can be effectively described in terms of spatial resolution, temporal resolution (sampling rate), and portability. The most widely used functional neuroimaging techniques are EEG and fMRI. EEG scanners consist of arrays of electrodes placed directly on the user's scalp. They are extremely portable and have high temporal resolution. However, such scanners can only detect neural activity after first being filtered by the skull, leading to a major decrease in spatial resolution. Conversely, fMRI is based on electromagnetic waves which easily penetrate the skull. Therefore, it has high spatial resolution, but is lacking in most other aspects. It is hundreds of times more expensive than EEG, it is thousands of times slower than EEG, and it is virtually unmovable by its user.

@@

Besides multi-modal functional neuroimaging, investigators have also proposed uni-modal synthesis as a method of augmenting neuroimaging techniques. This task consists of conditionally synthetizing novel data based on previous data collected through the same modality. This method has been found to increase performance in downstream classifiers based on fMRI data.

In addition to uni-modal synthesis, investigators have also proposed cross-modal synthesis for augmenting neuroimaging techniques. This task consists of reconstructing data collected through a target modality based on data collected through a distinct source modality, by exploiting the inherent relation between them. In a setting where both source and target modalities are available, this task might have limited utility. However, the value of cross-modal synthesis comes from settings where only one modality is available. In such settings, existing data can be used to approximate data collected through an unavailable modality. Unfortunately, this approach has largely been used for augmenting structural neuroimaging, with limited analogous efforts for functional neuroimaging.

@@

First, if simultaneous multi-modal recordings are available, supervised learning can be used to model the mapping between the two domains based on paired samples. Unfortunately, the constraint of physical compatibility greatly restricts the pairs of modalities which can be investigated. For instance, EEG scanners have to be specifically manufactured with MR-compatible materials, in order to avoid physical injury during EEG-fMRI operation. Fortunately, the development of multi-modal methods of data integration has highlighted multiple candidate pairs of modalities for simultaneous recording.

Second, if simultaneous multi-modal recordings are not available, but separate recordings of analogous mental states with different modalities are available (i.e. two recording sessions based on the same task but using different techniques), semi-supervised learning can be used to model the mapping through a task-related signal. This is analogous to the task of machine translation, which is based on a cross-lingual signal.

Third, if few such recordings exist, unsupervised learning can be used to learn the mapping by aligning latent spaces associated with the two modalities. This is analogous to machine translation where no paired multilingual corpora exist.

The suitability of EEG-fMRI as a candidate pair for cross-modal image synthesis is supported by the established practices of multi-modal neuroimaging using EEG-fMRI. These practices are the result of biophysical compatibility and the presence of complementary qualities.

Not only are EEG and fMRI compatible and complementary, but their results are also strongly correlated. The correlation occurs because both techniques are designed to detect neural activity.

>>>>

Neural activity is central to investigating cognition. In cognitive neuroscience, it has been employed to study the neural correlates of numerous cognitive functions, such as perception, attention, and memory. In cognitive modelling, neural activity has been used to inform the development of a wide range of cognitive architectures and models. In clinical neuroscience, it has been employed for investigating, diagnosing, and ameliorating a wide range of neurological conditions.

The increased relevance of neural activity in cognitive science prompted the development of a large number of functional neuroimaging techniques. These techniques can be effectively described in terms of spatial resolution, temporal resolution (sampling rate), and portability. The most widely used functional neuroimaging techniques are EEG and fMRI. EEG scanners consist of arrays of electrodes placed directly on the user's scalp. They are extremely portable and have high temporal resolution. However, such scanners can only detect neural activity after first being filtered by the skull, leading to a major decrease in spatial resolution. Conversely, fMRI is based on electromagnetic waves which easily penetrate the skull. Therefore, it has high spatial resolution, but is lacking in most other aspects. Compared to EEG, it is hundreds of times more expensive and thousands of times slower. Moreover, it is virtually unmovable by its user.

@@

Besides multi-modal functional neuroimaging, investigators have also proposed uni-modal synthesis as a method of augmenting neuroimaging techniques. This consists of conditionally synthetizing novel data based on previous data collected through the same modality. This method has been found to increase performance in downstream classifiers based on fMRI data.

In addition to uni-modal synthesis, investigators have also proposed cross-modal synthesis for augmenting neuroimaging techniques. This consists of reconstructing data collected through a target modality based on data collected through a distinct source modality, by exploiting the inherent relation between them. In a setting where both source and target modalities are available, this task might have limited utility. However, the value of cross-modal synthesis comes from settings where only one modality is available. In such settings, existing data can be used to approximate data collected through an unavailable modality. Unfortunately, this approach has largely been used for augmenting structural neuroimaging, with limited analogous efforts for functional neuroimaging.

@@

First, if simultaneous multi-modal recordings are available, supervised learning can be employed to model the mapping between the two domains based on paired samples. Unfortunately, the constraint of physical compatibility greatly restricts the pairs of modalities which can be investigated. For instance, MR-compatible EEG scanners have to be manufactured with special materials, in order to avoid physical injury during EEG-fMRI operation. Fortunately, the development of multi-modal methods of data integration has highlighted multiple candidate pairs of modalities for simultaneous recording.

Second, if simultaneous multi-modal recordings are not available, but separate recordings of similar mental states with different modalities are available (i.e. two separate recording sessions based on the same task but using different techniques), semi-supervised learning can be used to model the mapping through a task-related signal. This is analogous to the task of machine translation, which is often based on sample pairing through a cross-lingual signal.

Third, if the quantitiy of such recordings is limited, unsupervised learning can be used to learn the mapping by aligning latent spaces associated with the two modalities. This is analogous to machine translation where no paired multilingual corpora exist.

The following sections focus on supervised cross-modal synthesis, as the setup it requires is the most straight-forward.

\section{EEG-fMRI}

The suitability of EEG-fMRI as a candidate pair for cross-modal image synthesis is supported by the established practices of multi-modal neuroimaging using EEG-fMRI. These practices are the result of physical compatibility and the presence of complementary qualities.

Not only are EEG and fMRI compatible and complementary, but their results are also strongly correlated. The correlation occurs because both techniques can be used to record the same neural activity, albeit in different ways.
